A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2024-01-15 paper.Rmd
sec 4 done

In diagnosing a model fit using the residuals, we are generally interested in testing whether "*the regression model is correctly specified*" ($H_0$) against the broad alternative  "*the regression model is misspecified*" ($H_a$).

However, it is practically impossible to test this broad $H_0$ with conventional tests, because they need specific structure causing the departure to be quantifiable in order to be computable.  For example, the RESET test for detecting non-linear departures is formulated by fitting $y = \tau_0 + \sum_{i=1}^{p}\tau_px_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)$ in order to test $H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0$ against $H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0$. 
Similarly, the BP test is designed to specifically test $H_0:$ *error variances are all equal* ($\zeta_i=0 \text{ for } i=1,..,p$) versus the alternative $H_a:$ *that the error variances are a multiplicative function of one or more variables* ($\text{at least one } \zeta_i\neq 0$) from $e^2 = \zeta_0 + \sum_{i=1}^{p}\zeta_i x_i + u, ~ u\sim N(0,\sigma_u^2)$.

@@

The power of a conventional hypothesis test is affected by both the true parameter $\boldsymbol{\theta}$ and the sample size $n$. These two can be quantified in terms of effect size $E$ to measure the strength of the residual departures from the model assumptions. Details about the calculation of effect size are provided in Section \ref{effect-size} after the introduction of the simulation model used in our experiment. The theoretical power of a test is sometimes not a trivial solution, but it can be estimated if the data generating process is known. We use a predefined model to generate a large set of simulated data under different effect sizes, and record if the conventional test rejects $H_0$. The probability of the conventional test rejects $H_0$ is then fitted by a logistic regression formulated as

@@

The power of a visual test on the other hand, may additionally depend on the ability of the particular participant, as the skill of each individual may affect the number of observers who identify the data plot from the lineup [@majumder2013validation]. To address this issue, @majumder2013validation models the probability of participant $j$ correctly picking the data plot from lineup $l$ using a mixed-effect logistic regression, with participants treated as random effect. Then, the estimated power of a visual test evaluated by a single participant is the predicted value obtained from the mixed effects model. However, this mixed effects model does not work with scenario where participants are asked to select one or more most different plots. In this scenario, having the probability of a participant $j$ correctly picking the data plot from a lineup $l$ is insufficient to determine the power of a visual test because it does not provide information about the number of selections made by the participant for the calculation of the $p$-value. Therefore, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. This assumption essentially averages out the subject ability and helps to simplify the model structure, thereby obviating a costly large-scale experiment to estimate complex covariance matrices. The same model given in Equation \ref{eq:logistic-regression-1-1} is applied to model the power of a visual test.

@@

```{r different-shape-of-heter, fig.pos="!h", fig.height = 2.67, fig.cap = 'Heteroskedasticity forms used in the experiment. Three different shapes ($a = -1, 0, 1$) are used in the experiment to create left-triangle, "butterfly" and "right-triangle" shapes, respectively.'}

@@

Fitted values are a function of the independent variables, and the distribution of the observed values affects the distribution of the fitted values. In the best case scenario the fitted values will have a uniform distribution, which means that there is even coverage of possible observed values across all of the predictors. This is not always present in the collected data. Sometimes the fitted values are discrete because one or more predictors were measured discretely.  It is also common to see a skewed distribution of fitted values if one or more of the predictors has a skewed distribution. This latter problem is usually corrected before modelling, using a variable transformation. Our simulation assess this by using four different distributions to represent fitted values, constructed by different sampling of the predictor, including $U(-1, 1)$ (uniform), $N(0, 0.3^2)$ (normal), $\text{lognormal}(0, 0.6^2)/3$ (skewed) and $U\{-1, 1\}$ (discrete).

Figure \ref{fig:different-dist} shows the non-linear pattern, a "U" shape, with the different fitted value distributions. We would expect that structure in residual plots would be easier to perceive when the fitted values are uniformly distributed.

@@

The lineups are allocated to participants in a manner that covers the experimental design and is relatively uniform. To manage this, we use effect size to measure the signal strength, which helps in assigning a set of lineups with a range of difficulties to each participant.

@@

To compute the effect size for each lineup we simulate a sufficient large number of samples from the same model with the number of observations $n$ fixed for each sample. We then compute the effect size for each sample and take the average as the final value. This ensures lineups constructed with the same experimental factors will share the same effect size.

>>>>

In diagnosing a model fit using the residuals, we are generally interested in testing whether "*the regression model is correctly specified*" ($H_0$) against the broad alternative  "*the regression model is misspecified*" ($H_a$). However, it is practically impossible to test this broad $H_0$ with conventional tests, because they need specific structure causing the departure to be quantifiable in order to be computable.  For example, the RESET test for detecting non-linear departures is formulated by fitting $y = \tau_0 + \sum_{i=1}^{p}\tau_px_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)$ in order to test $H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0$ against $H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0$. 
Similarly, the BP test is designed to specifically test $H_0:$ *error variances are all equal* ($\zeta_i=0 \text{ for } i=1,..,p$) versus the alternative $H_a:$ *that the error variances are a multiplicative function of one or more variables* ($\text{at least one } \zeta_i\neq 0$) from $e^2 = \zeta_0 + \sum_{i=1}^{p}\zeta_i x_i + u, ~ u\sim N(0,\sigma_u^2)$.

@@

The power of a conventional hypothesis test is affected by both the true parameters $\boldsymbol{\theta}$ and the sample size $n$. These two can be quantified in terms of effect size $E$ to measure the strength of the residual departures from the model assumptions. Details about the calculation of effect size are provided in Section \ref{effect-size} after the introduction of the simulation model used in our experiment. The theoretical power of a test is sometimes not a trivial solution, but it can be estimated if the data generating process is known. We use a predefined model to generate a large set of simulated data under different effect sizes, and record if the conventional test rejects $H_0$. The probability of the conventional test rejects $H_0$ is then fitted by a logistic regression formulated as

@@

The power of a visual test on the other hand, may additionally depend on the ability of the particular participant, as the skill of each individual may affect the number of observers who identify the data plot from the lineup [@majumder2013validation]. To address this issue, @majumder2013validation models the probability of participant $j$ correctly picking the data plot from lineup $l$ using a mixed-effect logistic regression, with participants treated as random effects. Then, the estimated power of a visual test evaluated by a single participant is the predicted value obtained from the mixed effects model. However, this mixed effects model does not work with scenario where participants are asked to select one or more most different plots. In this scenario, having the probability of a participant $j$ correctly picking the data plot from a lineup $l$ is insufficient to determine the power of a visual test because it does not provide information about the number of selections made by the participant for the calculation of the $p$-value. Therefore, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. This assumption essentially averages out the subject ability and helps to simplify the model structure, thereby obviating a costly large-scale experiment to estimate complex covariance matrices. The same model given in Equation \ref{eq:logistic-regression-1-1} is applied to model the power of a visual test.

@@

```{r different-shape-of-heter, fig.pos="!h", fig.height = 2.67, fig.cap = 'Heteroskedasticity forms used in the experiment. Three different shapes ($a = -1, 0, 1$) are used in the experiment to create "left-triangle", "butterfly" and "right-triangle" shapes, respectively.'}

@@

Fitted values are a function of the independent variables (or predictors), and the distribution of the observed values affects the distribution of the fitted values. Ideally, we would want the fitted values to have a uniform coverage across the range of observed values or have a uniform distribution across all of the predictors. This is not always present in the collected data. Sometimes the fitted values are discrete because one or more predictors were measured discretely.  It is also common to see a skewed distribution of fitted values if one or more of the predictors has a skewed distribution. This latter problem is usually corrected before modelling, using a variable transformation. Our simulation assess this by using four different distributions to represent fitted values, constructed by different sampling of the predictor, including $U(-1, 1)$ (uniform), $N(0, 0.3^2)$ (normal), $\text{lognormal}(0, 0.6^2)/3$ (skewed) and $U\{-1, 1\}$ (discrete).

Figure \ref{fig:different-dist} shows the non-linear pattern, a "U" shape, with the different fitted value distributions. We would expect that structure in residual plots would be easier to perceive when the predictors are uniformly distributed.

@@

The lineups are allocated to participants in a manner that uniformly covers the combination of experimental factors in Table \ref{tab:model-factor-table}. In addition, we use effect size to measure the signal strength, which helps in assigning a set of lineups with a range of difficulties to each participant.

@@

To compute the effect size for each lineup we simulate a sufficiently large number of samples from the same model with the number of observations $n$ fixed for each sample. We then compute the effect size for each sample and take the average as the final value. This ensures lineups constructed with the same experimental factors will share the same effect size.

### 2024-01-15 paper.tex
sec 4 done

In diagnosing a model fit using the residuals, we are generally
interested in testing whether ``\emph{the regression model is correctly
specified}'' (\(H_0\)) against the broad alternative ``\emph{the
regression model is misspecified}'' (\(H_a\)).

However, it is practically impossible to test this broad \(H_0\) with
conventional tests, because they need specific structure causing the
departure to be quantifiable in order to be computable. For example, the
RESET test for detecting non-linear departures is formulated by fitting
\(y = \tau_0 + \sum_{i=1}^{p}\tau_px_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)\)
in order to test \(H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0\) against
\(H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0\).
Similarly, the BP test is designed to specifically test \(H_0:\)
\emph{error variances are all equal}
(\(\zeta_i=0 \text{ for } i=1,..,p\)) versus the alternative \(H_a:\)
\emph{that the error variances are a multiplicative function of one or
more variables} (\(\text{at least one } \zeta_i\neq 0\)) from
\(e^2 = \zeta_0 + \sum_{i=1}^{p}\zeta_i x_i + u, ~ u\sim N(0,\sigma_u^2)\).

@@

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
calculation of effect size are provided in Section \ref{effect-size}
after the introduction of the simulation model used in our experiment.
The theoretical power of a test is sometimes not a trivial solution, but
it can be estimated if the data generating process is known. We use a
predefined model to generate a large set of simulated data under
different effect sizes, and record if the conventional test rejects
\(H_0\). The probability of the conventional test rejects \(H_0\) is
then fitted by a logistic regression formulated as

@@

The power of a visual test on the other hand, may additionally depend on
the ability of the particular participant, as the skill of each
individual may affect the number of observers who identify the data plot
from the lineup \citep{majumder2013validation}. To address this issue,
\citet{majumder2013validation} models the probability of participant
\(j\) correctly picking the data plot from lineup \(l\) using a
mixed-effect logistic regression, with participants treated as random
effect. Then, the estimated power of a visual test evaluated by a single
participant is the predicted value obtained from the mixed effects
model. However, this mixed effects model does not work with scenario
where participants are asked to select one or more most different plots.
In this scenario, having the probability of a participant \(j\)
correctly picking the data plot from a lineup \(l\) is insufficient to
determine the power of a visual test because it does not provide
information about the number of selections made by the participant for
the calculation of the \(p\)-value. Therefore, we directly estimate the
probability of a lineup being rejected by assuming that individual skill
has negligible effect on the variation of the power. This assumption
essentially averages out the subject ability and helps to simplify the
model structure, thereby obviating a costly large-scale experiment to
estimate complex covariance matrices. The same model given in Equation
\ref{eq:logistic-regression-1-1} is applied to model the power of a
visual test.

@@

\caption{Heteroskedasticity forms used in the experiment. Three different shapes ($a = -1, 0, 1$) are used in the experiment to create left-triangle, "butterfly" and "right-triangle" shapes, respectively.}\label{fig:different-shape-of-heter}
\end{figure}

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. It is
also common to see a skewed distribution of fitted values if one or more
of the predictors has a skewed distribution. This latter problem is
usually corrected before modelling, using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values, constructed by different sampling of the
predictor, including \(U(-1, 1)\) (uniform), \(N(0, 0.3^2)\) (normal),
\(\text{lognormal}(0, 0.6^2)/3\) (skewed) and \(U\{-1, 1\}\) (discrete).

Figure \ref{fig:different-dist} shows the non-linear pattern, a ``U''
shape, with the different fitted value distributions. We would expect
that structure in residual plots would be easier to perceive when the
fitted values are uniformly distributed.

@@

To compute the effect size for each lineup we simulate a sufficient
large number of samples from the same model with the number of
observations \(n\) fixed for each sample. We then compute the effect
size for each sample and take the average as the final value. This
ensures lineups constructed with the same experimental factors will
share the same effect size.

>>>>

In diagnosing a model fit using the residuals, we are generally
interested in testing whether ``\emph{the regression model is correctly
specified}'' (\(H_0\)) against the broad alternative ``\emph{the
regression model is misspecified}'' (\(H_a\)). However, it is
practically impossible to test this broad \(H_0\) with conventional
tests, because they need specific structure causing the departure to be
quantifiable in order to be computable. For example, the RESET test for
detecting non-linear departures is formulated by fitting
\(y = \tau_0 + \sum_{i=1}^{p}\tau_px_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)\)
in order to test \(H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0\) against
\(H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0\).
Similarly, the BP test is designed to specifically test \(H_0:\)
\emph{error variances are all equal}
(\(\zeta_i=0 \text{ for } i=1,..,p\)) versus the alternative \(H_a:\)
\emph{that the error variances are a multiplicative function of one or
more variables} (\(\text{at least one } \zeta_i\neq 0\)) from
\(e^2 = \zeta_0 + \sum_{i=1}^{p}\zeta_i x_i + u, ~ u\sim N(0,\sigma_u^2)\).

@@

The power of a conventional hypothesis test is affected by both the true
parameters \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
calculation of effect size are provided in Section \ref{effect-size}
after the introduction of the simulation model used in our experiment.
The theoretical power of a test is sometimes not a trivial solution, but
it can be estimated if the data generating process is known. We use a
predefined model to generate a large set of simulated data under
different effect sizes, and record if the conventional test rejects
\(H_0\). The probability of the conventional test rejects \(H_0\) is
then fitted by a logistic regression formulated as

@@

The power of a visual test on the other hand, may additionally depend on
the ability of the particular participant, as the skill of each
individual may affect the number of observers who identify the data plot
from the lineup \citep{majumder2013validation}. To address this issue,
\citet{majumder2013validation} models the probability of participant
\(j\) correctly picking the data plot from lineup \(l\) using a
mixed-effect logistic regression, with participants treated as random
effects. Then, the estimated power of a visual test evaluated by a
single participant is the predicted value obtained from the mixed
effects model. However, this mixed effects model does not work with
scenario where participants are asked to select one or more most
different plots. In this scenario, having the probability of a
participant \(j\) correctly picking the data plot from a lineup \(l\) is
insufficient to determine the power of a visual test because it does not
provide information about the number of selections made by the
participant for the calculation of the \(p\)-value. Therefore, we
directly estimate the probability of a lineup being rejected by assuming
that individual skill has negligible effect on the variation of the
power. This assumption essentially averages out the subject ability and
helps to simplify the model structure, thereby obviating a costly
large-scale experiment to estimate complex covariance matrices. The same
model given in Equation \ref{eq:logistic-regression-1-1} is applied to
model the power of a visual test.

@@

\caption{Heteroskedasticity forms used in the experiment. Three different shapes ($a = -1, 0, 1$) are used in the experiment to create "left-triangle", "butterfly" and "right-triangle" shapes, respectively.}\label{fig:different-shape-of-heter}
\end{figure}

@@

Fitted values are a function of the independent variables (or
predictors), and the distribution of the observed values affects the
distribution of the fitted values. Ideally, we would want the fitted
values to have a uniform coverage across the range of observed values or
have a uniform distribution across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. It is
also common to see a skewed distribution of fitted values if one or more
of the predictors has a skewed distribution. This latter problem is
usually corrected before modelling, using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values, constructed by different sampling of the
predictor, including \(U(-1, 1)\) (uniform), \(N(0, 0.3^2)\) (normal),
\(\text{lognormal}(0, 0.6^2)/3\) (skewed) and \(U\{-1, 1\}\) (discrete).

Figure \ref{fig:different-dist} shows the non-linear pattern, a ``U''
shape, with the different fitted value distributions. We would expect
that structure in residual plots would be easier to perceive when the
predictors are uniformly distributed.

@@

To compute the effect size for each lineup we simulate a sufficiently
large number of samples from the same model with the number of
observations \(n\) fixed for each sample. We then compute the effect
size for each sample and take the average as the final value. This
ensures lineups constructed with the same experimental factors will
share the same effect size.
